---
title: "Understanding Uncertainty in Ecological Forecasts"
author: "Tadhg Moore, Mary Lofton, Quinn Thomas, Cayelan Carey"
date: "`r Sys.Date()`"
output:
  github_document: default
---

```{r setup, include=FALSE}
#knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(eval = FALSE)
```

## Purpose of this R Markdown

This R Markdown contains code to reproduce the basic functionality of "Macrosystems EDDIE Module 6: Understanding Uncertainty in Ecological Forecasts" outside of R Shiny. The code can be used by students to better understand what is happening "under the hood" of the Shiny app, which can be found at the following link:  
https://macrosystemseddie.shinyapps.io/module6/. 
  
Alternatively, students can complete this version of the module instead of the Shiny app version.  
  
## Summary
**Ecological forecasting** is a tool that can be used for understanding and predicting changes in populations, communities, and ecosystems. Ecological forecasting is an emerging approach which provides an estimate of the future state of an ecological system with uncertainty, allowing society to prepare for changes in important ecosystem services. Forecast **uncertainty** is derived from multiple sources, including model parameters and driver data, among others. Knowing the uncertainty associated with a forecast enables forecast users to evaluate the forecast and make more informed decisions. Ecological forecasters develop and update forecasts using the iterative forecasting cycle, in which they make a hypothesis of how an ecological system works; embed their hypothesis in a model; and use the model to make a forecast of future conditions and quantify forecast uncertainty. There are a number of approaches that forecasters can use to reduce uncertainty, which will be explored in this module.  
  
This module will guide you through an exploration of the sources of uncertainty within an ecological forecast, how uncertainty can be quantified, and steps which can be taken to reduce the uncertainty in a forecast you develop for a lake ecosystem.  
  
## Learning Outcomes
1. Define ecological forecast uncertainty.  
2. Explore the contributions of different sources of uncertainty (e.g., model parameters, model driver data) to total forecast uncertainty. 
3. Understand how multiple sources of uncertainty are quantified. 
4. Identify ways in which uncertainty can be reduced within an ecological forecast. 
5. Describe how forecast horizon affects forecast uncertainty. 
6. Explain the importance of specifying uncertainty in ecological forecasts for forecast users and decision support. 
  
## Key Concepts

**What is ecological forecast uncertainty?**. 
  
Uncertainty emerges from some kind of error or imperfection in our knowledge and understanding of the ecological system being investigated.  
  
**Where does ecological forecast uncertainty come from?**
  
Uncertainty comes from natural variability in the environment and imperfect knowledge of an ecological system. When generating a forecast, uncertainty can come from the structure of the model used, the conditions the model is initialized with, and the data used to drive the model, among other sources.  
  
**Why is uncertainty important to quantify for an ecological forecast?**. 
  
Knowing the uncertainty in a forecast allows forecast users to make informed decisions based on the range of forecasted outcomes and prepare accordingly.  
  
## Overview
In this module, we will generate forecasts of lake water temperature for 1-7 days into the future. First, we will generate a **deterministic** forecast (with no uncertainty). This will involve the following steps:  
  
1. Read in and visualize data from Lake Barco, FL, USA. 
2. Read in and visualize an air temperature forecast for Lake Barco. 
3. Build a multiple linear regression forecast model. 
4. Generate a deterministic forecast (without uncertainty). 
  
Next, we will explore how to incorporate four different kinds of uncertainty that are commonly present in **probabilistic** forecasts: driver data uncertainty, parameter uncertainty, process uncertainty, and initial conditions uncertainty. We will generate forecasts that incorporate these sources of uncertainty one at a time to learn how each form of uncertainty is accounted for. This will involve the following steps:  
  
5. Generate a forecast with driver uncertainty. 
6. Generate a forecast with parameter uncertainty. 
7. Generate a forecast with process uncertainty. 
8. Generate a forecast with initial conditions uncertainty. 

Finally, we will put it all together to generate a forecast that incorporates all four sources of uncertainty. We will also explore the relative contributions of each source of uncertainty to total forecast uncertainty; this is known as **uncertainty partitioning**. This will involve the following steps:  
  
9. Generate a forecast incorporating all sources of uncertainty. 
10. Partition uncertainty.  
  
Example code is provided for steps 1-5, and you will be asked several short answer questions to interpret code output. Beginning in step 6, you will be guided to build on the module code by adjusting example code or developing your own code. The coding questions will build in difficulty and the amount of guidance provided will decrease as you progress from step 6 to step 10. Keep in mind that the example code is provided to help you, and use as much of it as you can in completing the questions embedded in steps 6-10. There are a total of 16 questions. Please see the module rubric for possible points per question and confirm with your instructor whether and how the module will be graded.  
  
## Set-up
We will install and load some packages that are needed to run the module code. If you do not currently have the packages below downloaded for RStudio, you will need to install them first using the `install.packages()` function.
```{r, message = FALSE}
# install.packages("tidyverse")
# install.packages("ncdf4")
# install.packages("lubridate")
# install.packages("RColorBrewer")
# install.packages("reshape")
# install.packages("ggthemes")
library(tidyverse)
library(lubridate)
library(RColorBrewer)
library(ggthemes)
```

### 1. Read in and visualize data from Lake Barco, FL, USA
Lake Barco is one of the lake sites in the U.S. National Ecological Observatory Network (NEON). Please refer to https://www.neonscience.org/field-sites/barc to learn more about this site.  
  
#### Water Temperature
Water temperature exerts a major influence on biological activity and growth, has an effect on water chemistry, can influence water quantity measurements, and governs the kinds of organisms that live in water bodies. 

Water temperature can have important effects on water quality, as changes in water temperature can directly or indirectly affect water quality variables such as dissolved oxygen, nutrient and heavy metal concentrations, and algae concentrations.
  
Freshwater ecosystems are currently experiencing a multitude of stressors such as land use change and climate change, which can affect water temperature.  
  
Being able to predict how water temperature may change in the short-term (up to 7-days into the future) can provide natural resource managers with critical information to take pro-active actions to prevent degradation of water quality.  
```{r}
# Read in and view lake data

lake <- "BARC" #this is the four-letter site code corresponding to Lake Barco

# Read in air temperature and water temperature data
lake_df <- read_csv("./data/BARC_airt_wtemp_celsius.csv")

# Set custom color palette for our plot - ooh we are fancy!! :-)
cols <- RColorBrewer::brewer.pal(8, "Dark2")

# Build time series plot of air temperature and water temperature
p1 <- ggplot() +
  geom_line(data = lake_df, aes(Date, airt, color = "Air temperature")) +
  geom_line(data = lake_df, aes(Date, wtemp, color = "Water temperature")) +
  scale_color_manual(values = cols[5:6], name = "") +
  ylab("Temperature (\u00B0C)") +
  xlab("Time") +
  guides(color = guide_legend(override.aes = list(size = 3))) +
  theme_bw(base_size = 18)

# Render plot - this should match the time series plot you see in the Shiny app in 
# Activity A, Objective 3: Build a water temperature model IF you selected Lake Barco as your site
p1
```
**Question 1** Describe observed patterns in air and water temperature at Lake Barco over the course of a year, including the range of observed temperatures and whether and how air and water temperature  appear to relate to each other.  
  
**Answer 1**
  
  
### 2. Read in and visualize an air temperature forecast for Lake Barco 
We expect that future air temperatures will affect future water temperatures, and so we will use air temperature forecasts to help create water temperature forecasts.  
  
We have obtained an air temperature forecast for Lake Barco from the U.S. National Oceanic and Atmospheric Administration Global Ensemble Forecast System (NOAA GEFS). 
  
NOAA GEFS forecasts are **ensemble forecasts**. Ensemble forecasts are generated by running a model many times with different conditions. For example, a weather model used for forecasting might be run many times using slightly different starting conditions, because it is difficult to observe the atmosphere perfectly and so we are not exactly sure what the current conditions are. All the model runs together are referred to as the **ensemble**. Each individual model run is referred to as an **ensemble member**. Forecasters typically generate tens to hundreds of ensemble members to build uncertainty into their forecasts.  

Here, we read in NOAA forecast data that has been wrangled into a suitable format for our exercise: a two-dimensional data frame containing a 7-day-ahead air temperature forecast with 30 ensemble members.  
  
```{r}
# Read in and view air temperature forecast

# Name our forecast date: we will be working with a NOAA forecast generated on 2020-09-25
fc_date = "2020-09-25"

noaa_fc <- read_csv("./data/BARC_airt_forecast_NOAA_GEFS.csv")

# Voila! We now have an object called "noaa_fc" which is a two-dimensional data frame containing a 7-day-ahead NOAA air temperature forecast. Let's look at "noaa_fc".

head(noaa_fc)
# Forecast_date: this is the date for which temperature is forecasted
# Forecast_variable: this is the variable being forecasted
# Ensemble_member: this is an identifier for each member of the 30-member ensemble
# value: this is the value of the forecasted variable (in our case, degrees Celsius)

# Now we will plot the NOAA forecast.

# Data wrangling of observed air temperature data at Lake Barco so we can plot observed and forecasted air temperature on one time series plot
lake_obs <- lake_df %>%
  filter(Date >= "2020-09-22" & Date <= "2020-10-02") %>%
  mutate(wtemp = ifelse(Date > fc_date, NA, wtemp),
         airt = ifelse(Date > fc_date, NA, airt))

# Defining another custom color palette :-)
l.cols <- RColorBrewer::brewer.pal(8, "Set2")[-c(1, 2)]

# Build plot
p2 <- ggplot() +
  geom_point(data = lake_obs, aes(Date, airt, color = "Observed air temp.")) +
  geom_line(data = noaa_fc, aes(Forecast_date, value, group = Ensemble_member, color = "Forecasted air temp."), alpha = 0.6)+
  geom_vline(xintercept = as.Date(fc_date), linetype = "dashed") +
  ylab("Temperature (\u00B0C)") +
  theme_bw(base_size = 12) +
  scale_color_manual(values = c("Observed air temp." = cols[1], "Forecasted air temp." = "gray"),
                     name = "",
                     guide = guide_legend(override.aes = list(
                       linetype = c("solid","blank"),
                       shape = c(NA,16))))

# Render plot - this should match the time series plot in Activity B, Objective 9 - Driver Uncertainty IF you selected Lake Barco as your site
p2
```
**Question 2** Why are there multiple gray lines on the plot above? What do they represent? Be as specific as you can.  
  
**Answer 2**
  
  
### 3. Build a multiple linear regression forecast model 
We will use observed water temperature and air temperature data to build a linear regression model to predict water temperature.  
  
#### Key Terms for Ecological Modeling and Forecasting 
  
**What is an ecological model?**

A model is a representation (physical, conceptual or mathematical) of a real ecological phenomenon that is difficult to observe directly (e.g., the abundance of foxes in a forest).

**What is a linear relationship?**

A linear relationship is a statistical term used to describe a straight-line relationship between two variables. Linear relationships can be expressed either in a graphical format where the variable and the constant are connected via a straight line or in a mathematical format where the independent variable is multiplied by the slope coefficient, added by a constant, which determines the dependent variable.
In our example, water temperature is the dependent variable and air temperature is the independent variable.

**What is model error?**

Model error is the difference between an observation and the estimated value from the model. In this module, we assess the fit of our model by calculating three metrics:  
  
1. Coefficient of determination ($R^2$) 
2. Mean bias 
3. Root mean square error (RMSE) 
  
See the code below to understand how these metrics are calculated. A lower RMSE and mean bias indicate better model performance, while a higher $R^2$ indicates better model performance.

**What is a parameter?**

A parameter is a value in a model that describes the rate of change in those states.

**What is a parameter distribution?**

A parameter distribution shows the possible values for a parameter value and how often they occur. For example, a normal distribution of a parameter has a bell-shaped curve, and the mean is the most likely value of that parameter.

```{r}
# Fit model

# Build data frame to fit model
model_data <- lake_df %>%
  mutate(wtemp_yday = lag(wtemp))

# Fit multiple linear regression model using both yesterday's water temperature and today's air temperature
fit <- lm(model_data$wtemp ~ model_data$airt + model_data$wtemp_yday)
fit.summ <- summary(fit)

# View model coefficients and save them for our forecasts later
coeffs <- round(fit$coefficients, 2)
coeffs

# View standard errors of estimated model coefficients and save them for
# our forecasts later
params.se <- fit.summ$coefficients[,2]
params.se

# Calculate model predictions
mod <- predict(fit, model_data)

# Assess model fit
r2 <- round(fit.summ$r.squared, 2) #R2
err <- mean(mod - model_data$wtemp, na.rm = TRUE) #mean bias
rmse <- round(sqrt(mean((mod - model_data$wtemp)^2, na.rm = TRUE)), 2) #RMSE

# Prepare data frames for plotting
lake_df2 <- lake_df %>%
  filter(Date > "2020-01-01")
  
pred <- tibble(Date = model_data$Date,
                        Model = mod) %>%
  filter(Date > "2020-01-01")

# Build plot of modeled and observed water temperature
p3 <- ggplot() +
    geom_point(data = lake_df2, aes(Date, wtemp, color = "Observed")) +
    geom_line(data = pred, aes(Date, Model, color = "Modeled")) +
    ylab("Temperature (\u00B0C)") +
    xlab("Time") +
    scale_color_manual(values = c( "Observed" = "black", "Modeled" = cols[6]),
                       name = "",
                       guide = guide_legend(override.aes = list(
                         linetype = c("solid","blank"),
                         shape = c(NA,16)))) +
    theme_bw(base_size = 12) 

# Render plot - this should match the plot you see in the Shiny app Activity A, Objective 5 - Improve Model for Forecasting IF you selected Lake Barco as your site
p3
```
**Question 3** Describe the structure of the multiple linear regression model in your own words. How is water temperature being predicted?  
  
**Answer 3**
  
  
**Question 4** Examine the values of the fitted model parameters (`coeffs`). What does this tell you about the relative importance of past water temperature and current air temperature in driving current water temperature? Examine the standard errors of the parameters (`params.se`). What do the standard errors tell you about how confident we are in the fitted model parameter values?
  
**Answer 4**
  
  
**Question 5** Assess the model fit. Examine the values of `r2`, `err`, and `rmse`, as well as the plot showing the model fit and observations. What does this tell you about model performance?  
  
**Answer 5** 
  
  
### 4. Generate a deterministic forecast (without uncertainty)
Now we will generate a deterministic forecast with our model. We will use **one** ensemble member from the NOAA GEFS air temperature forecast ensemble as input to our multiple linear regression model, thus producing a water temperature prediction for 1 to 7 days into the future with no uncertainty.
```{r}
# Run deterministic forecast

# Setting up an empty dataframe that we will fill with our water temperature predictions
n_members = 1 #set number of ensemble members; this is set to 1 because we are making a deterministic forecast

# Set up date vector
forecast_horizon = 7
fc_dates <- seq.Date(from = ymd(fc_date), to = ymd(fc_date)+7, by = "day")

# Pull current water temperature to be initial condition
curr_wt <- lake_df %>% 
  filter(Date == fc_date) %>%
  select(wtemp)

#build forecast data frame
forecast_deterministic <- tibble(Forecast_date = rep(fc_dates, times = n_members),
                            Ensemble_member = rep(1:n_members, each = length(fc_dates)),
                            Forecast_variable = "water_temperature",
                            value = NA) %>%
  # insert today's water temperature as initial condition
  mutate(value = ifelse(Forecast_date == fc_date, curr_wt$wtemp, NA)) 
                            
# Run forecast. Here, we loop through days into the future and generate predictions using yesterday's water temperature and today's air temperature
for(d in 2:length(fc_dates)) {
  
  #pull prediction dataframe for relevant date
    temp_pred <- forecast_deterministic %>%
      filter(Forecast_date == fc_dates[d])
    
  #pull driver data for relevant date; here we select only 1 ensemble member from the NOAA air temperature forecast
    temp_driv <- noaa_fc %>%
      filter(Forecast_date == fc_dates[d] & Ensemble_member == 1)
    
  #pull lagged water temp values
    temp_lag <- forecast_deterministic %>%
      filter(Forecast_date == fc_dates[d-1])
    
  #run model
    temp_pred$value <- temp_driv$value * coeffs[2] + temp_lag$value * coeffs[3] + coeffs[1]
    
  #insert values back into forecast dataframe
    forecast_deterministic <- forecast_deterministic %>%
      rows_update(temp_pred, by = c("Forecast_date","Ensemble_member","Forecast_variable"))
}

# Build plot
p4 <- ggplot() +
  geom_point(data = lake_obs, aes(Date, wtemp, color = "Observed water temp.")) +
  geom_line(data = forecast_deterministic, aes(Forecast_date, value, color = "Forecasted water temp.")) +
  geom_vline(xintercept = as.Date(fc_date), linetype = "dashed") +
  ylab("Temperature (\u00B0C)") +
  theme_bw(base_size = 12) +
  scale_color_manual(values = c("Forecasted water temp." = cols[4],"Observed water temp." = cols[2]),
                     name = "",
                     guide = guide_legend(override.aes = list(
                       linetype = c("solid","blank"),
                       shape = c(NA, 16))))

# Render plot - this should resemble the plot in the R Shiny app Activity B Overview, labeled "Water Temperature Forecast"; here, we are plotting the "Both" model, which uses both yesterday's water temperature and today's forecasted air temperature to forecast water temperature
p4
```
**Question 6** Compare the water temperature forecast plot above to the plot of the NOAA GEFS air temperature forecast. Does the forecasted water temperature track the forecasted air temperature, and why might this be?  
  
**Answer 6** 
  
  
#### What is wrong with deterministic forecasts?
Using a **deterministic** forecast (e.g. a forecast which is one single line, with no uncertainty) is guaranteed to be wrong, because it ignores the uncertainty that is inherently associated with the future.

There are many things which contribute to uncertainty when generating a forecast, and a forecast should represent the range of potential outcomes and the **likelihood** of such outcomes occurring.

Therefore, we need to generate a **probabilistic** forecast which represents both the range of outcomes and also the likelihood of each.  
  
### 5. Generate a forecast with driver uncertainty
As a first step towards developing a probabilistic forecast, we will generate a forecast that incorporates **driver uncertainty**. Driver uncertainty comes from inaccuracies in the forecasted variables used as inputs to the forecast model. The driver variable for our model is air temperature. To generate a forecast of future water temperature that incorporates driver uncertainty, we need to use all 30 members of the NOAA GEFS air temperature forecast ensemble and generate a water temperature forecast with each one. Together, these 30 water temperature forecasts, each generated using a different NOAA GEFS ensemble member, will comprise an ensemble forecast of water temperature that accounts for driver uncertainty.

```{r}
# Run forecast with driver data uncertainty

# Setting up an empty dataframe that we will fill with our water temperature predictions
n_members = 30 #set number of ensemble members; notice we now use the entire NOAA forecast ensemble with 30 members

# Set up date vector
forecast_horizon = 7
fc_dates <- seq.Date(from = ymd(fc_date), to = ymd(fc_date)+7, by = "day")

# Pull current water temperature to be initial condition
curr_wt <- lake_df %>% 
  filter(Date == fc_date) %>%
  select(wtemp)

#build forecast data frame
forecast_ensemble <- tibble(Forecast_date = rep(fc_dates, times = n_members),
                            Ensemble_member = rep(1:n_members, each = length(fc_dates)),
                            Forecast_variable = "water_temperature",
                            value = NA) %>%
  # insert today's water temperature as initial condition
  mutate(value = ifelse(Forecast_date == fc_date, curr_wt$wtemp, NA)) 
                            
# Run forecast. Here, instead of looping through days into the future, we are looping through ensemble members and indexing the previous row in the matrix when we need to grab the previous water temperature. So we end up with a forecast corresponding to every member of the 30-member NOAA ensemble forecast.
for(d in 2:length(fc_dates)) {
  
  #pull prediction dataframe for relevant date
    temp_pred <- forecast_ensemble %>%
      filter(Forecast_date == fc_dates[d])
    
  #pull driver ensemble for relevant date; here we are using all 30 NOAA ensemble members
    temp_driv <- noaa_fc %>%
      filter(Forecast_date == fc_dates[d])
    
  #pull lagged water temp values
    temp_lag <- forecast_ensemble %>%
      filter(Forecast_date == fc_dates[d-1])
    
  #run model
    temp_pred$value <- temp_driv$value * coeffs[2] + temp_lag$value * coeffs[3] + coeffs[1]
    
  #insert values back into forecast dataframe
    forecast_ensemble <- forecast_ensemble %>%
      rows_update(temp_pred, by = c("Forecast_date","Ensemble_member","Forecast_variable"))
}

# Build plot
p5 <- ggplot() +
  geom_point(data = lake_obs, aes(Date, wtemp, color = "Observed water temp.")) +
  geom_line(data = forecast_ensemble, aes(Forecast_date, value, color = "Forecasted water temp.", group = Ensemble_member), alpha = 0.6) +
  geom_vline(xintercept = as.Date(fc_date), linetype = "dashed") +
  ylab("Temperature (\u00B0C)") +
  theme_bw(base_size = 12) +
  scale_color_manual(values = c("Forecasted water temp." = cols[6],
                                "Observed water temp." = cols[2]),
                     name = "",
                     guide = guide_legend(override.aes = list(
                       linetype = c("solid","blank"),
                       shape = c(NA, 16))))

# Render plot - this should resemble the water temperature forecast plot in the R Shiny app, Activity B Objective 9 ("Both" model)
p5

# Calculate standard deviation of forecast to quantify uncertainty later
sd_df <- forecast_ensemble %>%
  group_by(Forecast_date) %>%
  summarize(sd = sd(value, na.rm = TRUE)) %>%
  add_column(label = "Driver")
```
**Question 7** Compare the plot above of the probabilistic forecast with driver uncertainty to the plot of the deterministic forecast. How do the differences in these two forecasts affect your understanding of what water temperatures are likely to be during the week of the forecast?  
  
**Answer 7** 
  
  
### 6. Generate a forecast with parameter uncertainty
**Parameter uncertainty** refers to the uncertainty in the model parameter values, which can be due to uncertainties in the data or the calibration process used.

With traditional modelling efforts, people general find one set of the 'best fit' parameters and use them to predict with their model. 

This method does not account for the uncertainty around the estimation of these parameters.  

There is often the possibility that different parameter sets can yield similar metrics of model performance, e.g., similar R-squared values.

Using parameter distributions allows for a better representation of the potential predicted outcomes, leading to better quantification of the uncertainty.  
  
Here, you will use the standard errors of the parameters we estimated to generate distributions around each model parameter. You will then use these parameter distributions to incorporate parameter uncertainty into your water temperature forecasts.

Our model has three parameters: $\beta_0$, the intercept of the linear regression, $\beta_1$, the coefficient on tomorrow's air temperature, and $\beta_2$, the coefficient on today's water temperature. 

$$WaterTemp_{t+1} = \beta_0 + \beta_1*AirTemp_{t+1} + \beta_2*WaterTemp_t$$
When we fit our model, we obtained an estimate of the error around the mean of each of these parameters, which is stored in the `params.se` object. So instead of thinking of parameters as fixed values, we can think of them as distributions (here, a normal distribution) with some mean $(\mu)$ and variance (here represented by standard deviation, or $\sigma$):

$$\beta_0 \sim {\mathrm Norm}(\mu, \sigma)$$

**Question 8**

Generate parameter distributions based on parameters estimates for linear model.  
  
**HINT** Examine the `coeffs` and `params.se` objects we created when we fit the multiple regression model above. Think about what information they contain and how you might use that information to generate a **parameter distribution**.
```{r}
coeffs
params.se
```
**HINT** Look at the help documentation for the `rnorm()` function, which can be used to generate draws from a normal distribution with a specified mean and standard deviation. Note that **standard error** is defined as the standard deviation on the mean.
```{r}
?rnorm

#create an example distribution
example_distribution <- tibble(draws = rnorm(n = 1000, mean = 0, sd = 1))

#plot an example distribution
ggplot(data = example_distribution, aes(draws))+
  geom_histogram()+
  theme_classic()
```
Use the `rnorm()` function and the information in the `coeffs` and `params.se` objects to generate parameter distributions for each of the parameters in the multiple regression model. 

**Answer 8**
```{r}

```

**Question 9**
Plot each of the parameter distributions you have created. 

**Answer 9**
```{r}

```

**Question 10**
Adjust the forecasting for-loop to incorporate parameter uncertainty into your forecasts. Edit the code below where you see "INSERT YOUR CODE HERE". 

**HINT** Similar to how we used each of the 30 NOAA GEFS ensemble members to generate 30 slightly different forecasts that made up an ensemble with driver uncertainty, you will need to use slightly different parameter values to generate multiple forecasts that together, make up an ensemble incorporating parameter uncertainty. So here, each of your water temperature forecast ensemble members will use the **same** NOAA forecast but will have **different** parameter values drawn from your parameter distributions. This allows us to quantify how much uncertainty is coming from our model parameters while holding all other sources of uncertainty constant. 

**Answer 10**
```{r}
# Run forecast with parameter uncertainty

# Setting up an empty dataframe that we will fill with our water temperature predictions
n_members = 30 #set number of ensemble members

# Set up date vector
forecast_horizon = 7
fc_dates <- seq.Date(from = ymd(fc_date), to = ymd(fc_date)+7, by = "day")

# Pull current water temperature to be initial condition
curr_wt <- lake_df %>% 
  filter(Date == fc_date) %>%
  select(wtemp)

#build forecast data frame
forecast_ensemble <- tibble(Forecast_date = rep(fc_dates, times = n_members),
                            Ensemble_member = rep(1:n_members, each = length(fc_dates)),
                            Forecast_variable = "water_temperature",
                            value = NA) %>%
  # insert today's water temperature as initial condition
  mutate(value = ifelse(Forecast_date == fc_date, curr_wt$wtemp, NA)) 
                            
# Run forecast. 
for(d in 2:length(fc_dates)) {
  
  #pull prediction dataframe for relevant date
    temp_pred <- forecast_ensemble %>%
      filter(Forecast_date == fc_dates[d])
    
  #pull in a single member of NOAA air temperature forecast ensemble
    temp_driv <- noaa_fc %>%
      filter(Forecast_date == fc_dates[d] & Ensemble_member == 1)
    
  #pull lagged water temp values
    temp_lag <- forecast_ensemble %>%
      filter(Forecast_date == fc_dates[d-1])
    
  #run model
    #[INSERT YOUR CODE HERE]
    
  #insert values back into forecast dataframe
    forecast_ensemble <- forecast_ensemble %>%
      rows_update(temp_pred, by = c("Forecast_date","Ensemble_member","Forecast_variable"))
}

# Build plot
p6 <- ggplot() +
  geom_point(data = lake_obs, aes(Date, wtemp, color = "Observed water temp.")) +
  geom_line(data = forecast_ensemble, aes(Forecast_date, value, color = "Forecasted water temp.", group = Ensemble_member), alpha = 0.6) +
  geom_vline(xintercept = as.Date(fc_date), linetype = "dashed") +
  ylab("Temperature (\u00B0C)") +
  theme_bw(base_size = 12) +
  scale_color_manual(values = c("Forecasted water temp." = cols[6],
                                "Observed water temp." = cols[2]),
                     name = "",
                     guide = guide_legend(override.aes = list(
                       linetype = c("solid","blank"),
                       shape = c(NA, 16))))

# Render plot - this should resemble the water temperature forecast plot in the R Shiny app, Activity B Objective 9 ("Both" model)
p6

# Calculate standard deviation of forecast to quantify uncertainty later
sd_param <- forecast_ensemble %>%
  group_by(Forecast_date) %>%
  summarize(sd = sd(value, na.rm = TRUE)) %>%
  add_column(label = "Parameter")
sd_df <- sd_df %>%
  left_join(., sd_param, by = c("Date"))
```
### 7. Generate a forecast with process uncertainty
**Process uncertainty** is uncertainty caused by our inability to model all processes as observed in the real world.

Our "simple" water temperature model uses today's water temperature and tomorrow's forecasted air temperature to forecast tomorrow's water temperature. 
  
$$WaterTemp_{t+1} = \beta_0 + \beta_1*AirTemp_{t+1} + \beta_2*WaterTemp_t$$

But we know that water temperature can be affected by other processes as well (such as rain, inflow streams to a lake, or water column mixing, to name a few) and that our model has simplified or ignored these. To account for the uncertainty these simplifications introduce to our model, we can add in process noise (W) to our model at each time step. In this model, water temperature tomorrow is equal to water temperature today plus air temperature tomorrow plus some noise (W),

$$WaterTemp_{t+1} = \beta_0 + \beta_1*AirTemp_{t+1} + \beta_2*WaterTemp_t + W$$
where process noise is equal to a random number drawn from a normal distribution with a mean of zero and a standard deviation ($\sigma$).

$$W \sim {\mathrm Norm}(0, \sigma)$$
To account for process uncertainty, we can run the model multiple times with random noise added to each model run. More noise is associated with high process uncertainty, and vice versa.  

**Question 11** Adjust the forecasting for-loop to incorporate process uncertainty into your forecasts. Edit the code below where you see "INSERT YOUR CODE HERE". Your forecast ensemble should have 30 members.

**HINT** Remember, we want to isolate the effect of process uncertainty on our forecast so we can quantify it. You should be sure to **only** incorporate process uncertainty below (i.e., driver data and parameter values should be constant across all ensemble members). Use the two equations above to guide you in building your forecast model equation that incorporates process uncertainty.

**Answer 11**
```{r}
# Run forecast with process uncertainty 

# Set value of noise that will be added to model
sigma <- 0.2 # Process Uncertainty Noise Std Dev.; this is your sigma

# Setting up an empty matrix that we will fill with our water temperature predictions
n_members = 30 #set number of ensemble members

# Set up date vector
forecast_horizon = 7
fc_dates <- seq.Date(from = ymd(fc_date), to = ymd(fc_date)+7, by = "day")

# Pull current water temperature to be initial condition
curr_wt <- lake_df %>% 
  filter(Date == fc_date) %>%
  select(wtemp)

#build forecast data frame
forecast_ensemble <- tibble(Forecast_date = rep(fc_dates, times = n_members),
                            Ensemble_member = rep(1:n_members, each = length(fc_dates)),
                            Forecast_variable = "water_temperature",
                            value = NA) %>%
  # insert today's water temperature as initial condition
  mutate(value = ifelse(Forecast_date == fc_date, curr_wt$wtemp, NA)) 
                            
# Run forecast. 
for(d in 2:length(fc_dates)) {
  
  #pull prediction dataframe for relevant date
    temp_pred <- forecast_ensemble %>%
      filter(Forecast_date == fc_dates[d])
    
  #pull in a single member of NOAA air temperature forecast ensemble
    temp_driv <- noaa_fc %>%
      filter(Forecast_date == fc_dates[d] & Ensemble_member == 1)
    
  #pull lagged water temp values
    temp_lag <- forecast_ensemble %>%
      filter(Forecast_date == fc_dates[d-1])
    
  #run model with process uncertainty added
    #[INSERT YOUR CODE HERE]
    
  #insert values back into forecast dataframe
    forecast_ensemble <- forecast_ensemble %>%
      rows_update(temp_pred, by = c("Forecast_date","Ensemble_member","Forecast_variable"))
}

# Build plot
p7 <- ggplot() +
  geom_point(data = lake_obs, aes(Date, wtemp, color = "Observed water temp.")) +
  geom_line(data = forecast_ensemble, aes(Forecast_date, value, color = "Forecasted water temp.", group = Ensemble_member), alpha = 0.6) +
  geom_vline(xintercept = as.Date(fc_date), linetype = "dashed") +
  ylab("Temperature (\u00B0C)") +
  theme_bw(base_size = 12) +
  scale_color_manual(values = c("Forecasted water temp." = cols[6],
                                "Observed water temp." = cols[2]),
                     name = "",
                     guide = guide_legend(override.aes = list(
                       linetype = c("solid","blank"),
                       shape = c(NA, 16))))

# Render plot - this should resemble the water temperature forecast plot in the R Shiny app, Activity B Objective 9 ("Both" model)
p7

# Calculate standard deviation of forecast to quantify uncertainty later
sd_proc <- forecast_ensemble %>%
  group_by(Forecast_date) %>%
  summarize(sd = sd(value, na.rm = TRUE)) %>%
  add_column(label = "Process")
sd_df <- sd_df %>%
  left_join(., sd_proc, by = c("Date"))
```
### 8. Generate a forecast with initial conditions uncertainty
Initial conditions uncertainty refers to uncertainty arising because the initial conditions are not precisely known or because the calculations cannot be performed with the precise initial conditions.

Even though we have measurements of water temperature from our lake, we know that water temperature varies throughout the day so this measurement might not capture exactly the temperature in our lake at this time. Additionally, there may be observation error in our temperature measurements.

To account for initial condition uncertainty we can generate a distribution around the initial condition and then run our model with slightly different initial conditions to account for this uncertainty.

The following questions guide you through the process of quantifying initial conditions uncertainty.

**Question 12** Generate a distribution of initial conditions for your forecast using the current water temperature (`curr_wt`) and a standard deviation of 0.1 degrees Celsius (`ic_uc`).

**Answer 12**
```{r}
# Generate an initial conditions distribution 
curr_wt <- lake_df %>% 
  filter(Date == fc_date) %>%
  select(wtemp)
ic_uc <- 0.1 

#INSERT YOUR CODE HERE
```

**Question 13** Plot the distribution around your initial condition.

**Answer 13**
```{r}

```

**Question 14** Generate a forecast that incorporates initial conditions uncertainty only, holding all other sources of uncertainty (driver, parameter, process) constant. Your ensemble should have 30 members, with each member having a slightly different initial conditions value. Be sure to both plot the forecast and save the standard deviation of the forecast to quantify initial conditions uncertainty. 

**HINT** Use code provided above as a starting point!

**Answer 14**
```{r}
# A. Randomly sample 30 values from our initial conditions distribution to use for our forecast


# B. Set up a dataframe that we will fill with our ensemble water temperature forecasts

# C. Fill the rows of the dataframe where Date == "2020-09-25" with the 30 values you randomly sampled from your initial conditions distribution


# D. Run forecast. This will be a for-loop looping through dates; see code above. Be sure you hold all sources of uncertainty besides initial conditions uncertainty constant; this means you will only use ONE ensemble member from the NOAA forecast, have fixed parameter values, and will not add process uncertainty.


# E. Build and render plot


# F. Calculate the standard deviation of each day of the forecast to quantify initial conditions uncertainty - we will use this information in the next section. Your final output should be a dateframe with the following columns:
# Forecast_date: yyyy-mm-dd
# sd: standard deviation across the 30 ensemble members for each forecast date
# Label: set to "Initial Conditions"


# G. Left-join your dataframe with the standard deviation of initial conditions to the dataframe of standard deviations with all forecasts included; example code is provided below

# sd_df <- sd_df %>%
#   left_join(.,[INSERT YOUR CODE HERE], by = "Forecast_date")

```
### 9. Generate a forecast incorporating all sources of uncertainty
To plot a forecast with all sources of uncertainty incorporated, we need to generate a forecast that incorporates driver, parameter, process, and initial conditions uncertainty.

**Question 15** Adjust the forecasting for-loop to incorporate all four sources of uncertainty (driver, parameter, process, initial conditions) into your forecasts. Edit the code below where you see "INSERT YOUR CODE HERE". Your forecast ensemble should have 30 members.  
  
**Answer 15**
```{r}
# A. Set up a dataframe that we will fill with our ensemble water temperature forecasts; see example code above


# B. Fill the rows of the dataframe where Date == "2020-09-25" with 30 values randomly sampled from your initial conditions distribution


# C. Run forecast. This will be a for-loop looping through dates; see code above. Be sure to include driver, process, initial condition, and parameter uncertainty


# D. Build and render a time series plot of your forecast; see example code above


```
### 10. Partition uncertainty 
If you have successfully calculated and saved the standard deviation from each of your forecasts incorporating a single source of forecast uncertainty above, you should be able to run the following code to see the relative contributions of each source of uncertainty to total forecast uncertainty.

**HINT** This code chunk requires there to be an object in your environment called `df.ic`, which you generated as part of the answer to Question 7. `df.ic` should contain the standard deviations of your forecast with initial conditions uncertainty and should follow a specific format described in Question 7. If you get an error, double-check that you have correctly built and saved `df.ic`.
```{r}
# Quantify uncertainty

# Set another custom plot color palette
cols2 <- ggthemes::ggthemes_data$colorblind$value

# Plot the contribution of each source of uncertainty to total forecast uncertainty
# Build plot
p9 <- ggplot() +
  geom_bar(data = sd_df, aes(Forecast_date, sd, fill = label), stat = "identity", position = "stack") +
  ylab("Standard Deviation (\u00B0C)") +
  scale_fill_manual(values = c("Process" = cols2[1], "Parameter" = cols2[2], "Initial Condition" = cols2[3],
                               "Driver" = cols2[4], "Total" = cols2[5])) +
  scale_x_date(date_breaks = "1 day", date_labels = "%b %d") +
  labs(fill = "Uncertainty") +
  theme_bw(base_size = 12)

# Render plot - this should resemble the uncertainty quantification plot in the R Shiny app, Activity C, Objective 10 ("Both" model)
p9
```
**Question 16 Optional Exercise** Parameter uncertainty is a substantial source of uncertainty in our forecast. To constrain our forecast uncertainty and potentially improve our forecast, we could try to reduce the uncertainty around our parameters by collecting more environmental data and improving the fit of our linear regression model. However, before we do this, we might want to know **which** parameter is contributing the most to parameter uncertainty. Below, develop code that partitions the contribution of **each individual parameter** to parameter uncertainty, and plot their relative contributions. Which parameter should we focus on improving estimates of to reduce total forecast uncertainty? Based on your findings, what recommendations might you make to a water manager who is interested in improving the performance of the water temperature forecast?

**Answer 16**
```{r}

```

Congratulations! You have quantified all the uncertainty. Now, have a nap. :-)  
  
## Submitting Your Assignment
If you click the "Knit" button in RStudio, this assignment will be knitted to a .md file which can be committed and pushed to GitHub. Before you do this, be sure to comment out the following line in the `{r setup}` chunk:  
`knitr::opts_chunk$set(eval = FALSE)`  
This will ensure that all your code chunks are evaluated and your instructor can see your work.
